Shapes analysis for time series.

### 2024-05-22 NEURIPS24/experiment.tex
update

    \item \textbf{Classification benchmark on regularly sampled datasets, \Cref{appendix: settings_classification}:} We compare performances of a kernel support vector machine (SVC) algorithm based on TS-LDDMM representation with 3 state-of-the-art classification methods from shape analysis on 15 shape-based datasets (7 univariates \& 8 multivariates). \textbf{Results} in \Cref{table:wewin} show that the TS-LDDMM-based method outperforms other methods (best performances over 13 datasets), making TS-LDDMM representation relevant for time series shape analysis.
\end{enumerate}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/ts_ddmm_shooting.pdf}
    \caption{TS-LDDMM}
    \label{fig:ts-lddmm shooting}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/lddmm_shooting.pdf}
    \caption{LDDMM}
    \label{fig:lddmm shooting}
  \end{subfigure}
  \caption{Analysis of the two principal components (PC) related to mice' respiratory cycles before exposure for TS-LDDMM \Cref{fig:ts-lddmm shooting}, and LDDMM \Cref{fig:lddmm shooting}.
  In both cases, I) displays PC densities according to mice genotype and II) displays  deformations of the reference graph $S_0$ along each PC.}
  \label{fig:exp1}
\end{figure*}

\vspace{-1ex}
\subsection{Interpretability: analysis of respiratory behavior in mice}
\label{section:interpretability}
\vspace{-1ex}
%We consider a dataset composed of mouse respiratory time series before and after a drug injection.
% A complete presentation of this dataset is given in \Cref{appendix:mouse_dataset}.
%  The mouse are divided in two groups depending on their genotypes : \textit{colq} and \textit{wt}.
%  We aim to study the difference of respiratory shapes according to the genotype.
%   That is why, TS-LDDMM \eqref{eq:general_optimization_problem} is applied to derive the features representations $(\mathbf{\alpha}_j)_{j\in[N]}\in (\Rset^2)^N$ related to $N$=\sam{fill} respiratory time series coming from 14 mouse (7 \textit{colq} and \textit{wt}) before drug injection.
%   Then, a Principal Components Analysis (PCA) is performed on $(\mathbf{\alpha}_j)_{j\in[N]}$

@@

\paragraph{Protocol.} For both experiments and representations (TS-LDDMM and LDDMM), we derive the reference respiratory cycle's graph $S_0$ and the representations $(\alpha_j)_{j\in[N]}$ related to $N$ respiratory cycles extracted according to the procedure \cite{germain2023unsupervised} by solving  \eqref{eq:general_optimization_problem}.
 Then, we perform a kernel PCA on the initial velocity field $\left(v_0(\alpha_j,S_0)\right)_{j\in[N]}\in \msv^{N_1}$ defined in \eqref{eq:def_v0} to breathing behavior.
  The first experiment includes $N_1 = 700$ respiratory cycles collected before exposure. The second experiment includes $N_2 = 1400$ respiratory cycles with 25\% (resp. 75\%) before (resp. after) exposure.
   \Cref{appendix: mice_exp_setting} describes methods settings.
    Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales. 
    \begin{table}[t]
      \centering
      \resizebox{\columnwidth}{!}{%
      \begin{tabular}{llrrrr}
        \toprule
         & \textbf{Dataset} & \textbf{Shape-FPCA (2024)} & \textbf{TCLR (2024)} & \textbf{LDDMM (2008)} & \textbf{TS-LDDMM (ours)} \\
        \midrule
        \multirow[c]{7}{*}{Univariate} & ArrowHead & 0.18 & 0.75 & \underline{0.84} & \textbf{0.91} \\
         & BME & 0.16 & \underline{1.00} & 0.82 & \textbf{1.00} \\
         & ECG200 & 0.40 & 0.67 & \textbf{0.81} & \underline{0.79} \\
         & FacesUCR & 0.08 & \underline{0.73} & 0.69 & \textbf{0.86} \\
         & GunPoint & 0.93 & \underline{0.97} & 0.83 & \textbf{1.00} \\
         & PhalangesOutlinesCorrect & 0.39 & \textbf{0.63} & \underline{0.53} & 0.52 \\
         & Trace & 0.55 & \underline{1.00} & 0.46 & \textbf{1.00} \\
        \cline{1-6}
        \multirow[c]{8}{*}{Multivariate} & ArticularyWordRecognition & -- & -- & \underline{0.98} & \textbf{1.00} \\
         & Cricket & -- & -- & \underline{0.77} & \textbf{0.93} \\
         & ERing & -- & -- & \underline{0.95} & \textbf{0.98} \\
         & Handwriting & -- & -- & \underline{0.22} & \textbf{0.44} \\
         & Libras & -- & -- & \underline{0.56} & \textbf{0.60} \\
         & NATOPS & -- & -- & \underline{0.82} & \textbf{0.82} \\
         & RacketSports & -- & -- & \textbf{0.83} & \underline{0.79} \\
         & UWaveGestureLibrary & -- & -- & \underline{0.72} & \textbf{0.81} \\
        \bottomrule
        \end{tabular}
        %
      }
      \label{table:wewin}
      \caption{All the state-of-the-art methods \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} mentionned in related works are compared to TS-LDDMM in \Cref{appendix: settings_classification}.}

    \end{table}
    \vspace{-1ex}
    \vspace{-1ex}\begin{wrapfigure}{r}{0.5\textwidth}
      \centering
      \begin{subfigure}[b]{0.15\textwidth}
        \centering
        \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
        \caption{colq cycle}
        \label{fig:colq-cycle}
      \end{subfigure}
      \begin{subfigure}[b]{0.15\textwidth}
        \centering
        \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
        \caption{PC1 vs PC2}
        \label{fig: pcs-scatter}
      \end{subfigure}
      \begin{subfigure}[b]{0.15\textwidth}
        \centering
        \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
        \caption{wt cycle}
        \label{fig:wt-reference}
      \end{subfigure}
      \caption{\Cref{fig:colq-cycle} is a \textbf{colq} respiratory cycle. \Cref*{fig: pcs-scatter} displays individual mouse reference respiratory cycle in the TS-LDDMM PC1-PC2 coordinates. \Cref{fig:wt-reference} is a referent respiratory cycle of a \textbf{wt} mouse learned with TS-LDDMM. }
      \vspace{-1em}
    \end{wrapfigure}
\paragraph{Breathing behaviors before exposure.}
 We focus on the analysis of the two first Principal Components (PC) for TS-LDDMM (\Cref{fig:ts-lddmm shooting}), and LDDMM (\Cref{fig:lddmm shooting}).
  A respiratory cycle is an inspiration (positive flow) followed by an expiration (negative flow); \Cref{fig:wt-reference} displays an example of \textbf{wt} mouse respiratory cycle.
   \Cref{fig:exp1} shows that principal components learned with TS-LDMM lead to deformations that remain respiratory cycles, while deformations learned with LDDMM are challenging to interpret as respiratory cycles.
   The LDDMM velocity field kernel is a Gaussian anisotropic kernel that accounts for time and space scales; however, the entanglement of time and space dimensions in the kernel does not guarantee the graph structure, and it makes the convergence of the method difficult (relative varifold loss error: TS-LDDMM: 0.06, LDDMM: 0.11).
    Regarding TS-LDDMM, \Cref{fig:ts-lddmm shooting} shows its principal components refer to different types of deformations.
     By interpreting \Cref{fig:ts-lddmm shooting}: PC1 accounts for time warping, PC2 expresses the trade-off between inspiration and expiration duration.
      Compared to \textbf{wt} mice, the distribution of \textbf{colq} mice feature along the PC1 axis has a heavy left tail, and the associated deformation (-2 $\sigma_{\text{PC}}$) shows an inspiration with two peaks.
       \Cref{fig:colq-cycle} shows an example of such respiratory cycles, which may be caused by motor impairment due to enzyme deficiency \cite{germain2023unsupervised}.
        Finally, \Cref{fig: pcs-scatter} shows that PC1 and PC2 capture the main differences between the two groups as their respective reference graphs $S^j$ are located in different parts of the space.

>>>>

    \item \textbf{Classification benchmark on regularly sampled datasets, \Cref{appendix: settings_classification}:} We compare performances of a kernel support vector machine (SVC) algorithm based on TS-LDDMM representation with 3 state-of-the-art classification methods from shape analysis on 15 shape-based datasets (7 univariates \& 8 multivariates). \textbf{Results} show that the TS-LDDMM-based method outperforms other methods (best performances over 13 datasets), making TS-LDDMM representation relevant for time series shape analysis.
\end{enumerate}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/ts_ddmm_shooting.pdf}
    \caption{TS-LDDMM}
    \label{fig:ts-lddmm shooting}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/lddmm_shooting.pdf}
    \caption{LDDMM}
    \label{fig:lddmm shooting}
  \end{subfigure}
  \caption{Analysis of the two principal components (PC) related to mice' respiratory cycles before exposure for TS-LDDMM \Cref{fig:ts-lddmm shooting}, and LDDMM \Cref{fig:lddmm shooting}.
  In both cases, I) displays PC densities according to mice genotype and II) displays  deformations of the reference graph $S_0$ along each PC.}
  \label{fig:exp1}
\end{figure*}

\vspace{-1ex}
\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_cycle_example.jpeg}
    \caption{colq cycle}
    \label{fig:colq-cycle}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_scatter.jpeg}
    \caption{PC1 vs PC2}
    \label{fig: pcs-scatter}
  \end{subfigure}
  \begin{subfigure}[b]{0.15\textwidth}
    \centering
    \includegraphics[width = \textwidth]{pictures/exp_1_wt.jpeg}
    \caption{wt cycle}
    \label{fig:wt-reference}
  \end{subfigure}
  \caption{\Cref{fig:colq-cycle} is a \textbf{colq} respiratory cycle. \Cref*{fig: pcs-scatter} displays individual mouse reference respiratory cycle in the TS-LDDMM PC1-PC2 coordinates. \Cref{fig:wt-reference} is a referent respiratory cycle of a \textbf{wt} mouse learned with TS-LDDMM. }
  \vspace{-1em}
\end{figure*}
\vspace{-1ex}
\subsection{Interpretability: analysis of respiratory behavior in mice}
\label{section:interpretability}
\vspace{-1ex}
%We consider a dataset composed of mouse respiratory time series before and after a drug injection.
% A complete presentation of this dataset is given in \Cref{appendix:mouse_dataset}.
%  The mouse are divided in two groups depending on their genotypes : \textit{colq} and \textit{wt}.
%  We aim to study the difference of respiratory shapes according to the genotype.
%   That is why, TS-LDDMM \eqref{eq:general_optimization_problem} is applied to derive the features representations $(\mathbf{\alpha}_j)_{j\in[N]}\in (\Rset^2)^N$ related to $N$=\sam{fill} respiratory time series coming from 14 mouse (7 \textit{colq} and \textit{wt}) before drug injection.
%   Then, a Principal Components Analysis (PCA) is performed on $(\mathbf{\alpha}_j)_{j\in[N]}$

@@

\paragraph{Protocol.} For both experiments and representations (TS-LDDMM and LDDMM), we derive the reference respiratory cycle's graph $S_0$ and the representations $(\alpha_j)_{j\in[N]}$ related to $N$ respiratory cycles extracted according to the procedure \cite{germain2023unsupervised} by solving  \eqref{eq:general_optimization_problem}.
 Then, we perform a kernel PCA on the initial velocity field $\left(v_0(\alpha_j,S_0)\right)_{j\in[N]}\in \msv^{N_1}$ defined in \eqref{eq:def_v0} to breathing behavior.
  The first experiment includes $N_1 = 700$ respiratory cycles collected before exposure. The second experiment includes $N_2 = 1400$ respiratory cycles with 25\% (resp. 75\%) before (resp. after) exposure.
   \Cref{appendix: mice_exp_setting} describes methods settings.
    Essentially, varifold losses are identical for both methods, and the velocity field kernels are set to encompass time and space scales.

    \vspace{-1ex}
\paragraph{Breathing behaviors before exposure.}
 We focus on the analysis of the two first Principal Components (PC) for TS-LDDMM (\Cref{fig:ts-lddmm shooting}), and LDDMM (\Cref{fig:lddmm shooting}).
  A respiratory cycle is an inspiration (positive flow) followed by an expiration (negative flow); \Cref{fig:wt-reference} displays an example of \textbf{wt} mouse respiratory cycle.
   \Cref{fig:exp1} shows that principal components learned with TS-LDMM lead to deformations that remain respiratory cycles, while deformations learned with LDDMM are challenging to interpret as respiratory cycles.
   The LDDMM velocity field kernel is a Gaussian anisotropic kernel that accounts for time and space scales; however, the entanglement of time and space dimensions in the kernel does not guarantee the graph structure, and it makes the convergence of the method difficult (relative varifold loss error: TS-LDDMM: 0.06, LDDMM: 0.11).
    Regarding TS-LDDMM, \Cref{fig:ts-lddmm shooting} shows its principal components refer to different types of deformations.
     By interpreting \Cref{fig:ts-lddmm shooting}: PC1 accounts for time warping, PC2 expresses the trade-off between inspiration and expiration duration.
      Compared to \textbf{wt} mice, the distribution of \textbf{colq} mice feature along the PC1 axis has a heavy left tail, and the associated deformation (-2 $\sigma_{\text{PC}}$) shows an inspiration with two peaks.
       \Cref{fig:colq-cycle} shows an example of such respiratory cycles, which may be caused by motor impairment due to enzyme deficiency \cite{germain2023unsupervised}.
        Finally, \Cref{fig: pcs-scatter} shows that PC1 and PC2 capture the main differences between the two groups as their respective reference graphs $S^j$ are located in different parts of the space.

### 2024-05-22 NEURIPS24/paper.tex
update

 %LDDMM framework is a relevant shape analysis tool to represent curves as depicted in \cite{glaunes2008large}. However, graphs of time series are a well-structured type of curve due to the inclusion of the temporal dimension that requires specific care (\Cref{fig:diffeo}).
 %In a similar vein, Qiu \textit{et al} \cite{qiu2009time} proposes a method for tracking anatomical shape changes in serial images using LDDMM. They include temporal evolution, but not for the same purpose: the aim is to perform longitudinal modeling of brain images.
 Qiu \textit{et al} \cite{qiu2009time} proposes a method for tracking anatomical shape changes in serial images using LDDMM. They include temporal evolution, but not for the same purpose: the aim is to perform longitudinal modeling of brain images.
 Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} address the representation of curves with the Square-Root Velocity (SRV) representation.
 However, the SRV representation is applied after reparametrization of the temporal dimension of the unit length segment.
 Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
 Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
 Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.

@@

 %All these state-of-the-art methods previously mentionned \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} are compared to TS-LDDMM in \Cref{appendix: settings_classification}.
 \vspace{-1ex}
  \section{Limitations and conclusion}
  \vspace{-1ex}
  \label{sec:limitations}
  %\paragraph*{Limitations}
  This paper proposes a feature representation method, TS-LDDMM, designed for 
shape comparison on homogeneous time series datasets. We show on a real dataset 
its ability to study, with high interpretability, the inter-individual shape 
variability. As an unsupervised approach, it is user-friendly and enables knowledge 
transfer for different supervised tasks such as classification. Although TS-LDDMM 
is already competitive for classification, its performances can be leveraged on 
more heterogeneous datasets using a hierarchical clustering extension, which is relegated for future work. 
%   While TS-LDDMM performs well on shape-based datasets, it assumes a certain degree of homogeneity with the existence of a reference time series graph$\msg_0$.
%  Extending TS-LDDMM to more heterogeneous datasets using clustering is ongoing work.
 TS-LDDMM employs kernel computations, which require specific libraries (e.g., KeOps \cite{charlier2021kernel}) to be efficient and scalable.
 However, in our experiments, the time complexity of TS-LDDMM is comparable to that of competitors.
 It is clear that TS-LDDMM needs to be extended to handle very large datasets with high-dimensional time series (such as videos).
 Additionally, TS-LDDMM requires tuning several hyperparameters, though this is a common requirement among competitors \cite{glaunes2008large, oh2024stable, wu2024shape, heo2024logistic}.
 In future work, adaptive methods are expected to be developed to provide a more user-friendly interface.
  %\color{red} rappeler les limites de la méthodes : petite dimension à cause des kernels, tuner les hyper paramètres correctement, version adaptative}
% \paragraph*{Conclusion}
% \vspace{-1ex}

>>>>

 LDDMM framework is a relevant shape analysis tool to represent curves as depicted in \cite{glaunes2008large}. However, graphs of time series are a well-structured type of curve due to the inclusion of the temporal dimension that requires specific care (\Cref{fig:diffeo}).
 In a similar vein, Qiu \textit{et al} \cite{qiu2009time} proposes a method for tracking anatomical shape changes in serial images using LDDMM. They include temporal evolution, but not for the same purpose: the aim is to perform longitudinal modeling of brain images.

 Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} address the representation of curves with the Square-Root Velocity (SRV) representation.
 However, the SRV representation is applied after reparametrization of the temporal dimension of the unit length segment.
 Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
 Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
 Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.

@@

 All these state-of-the-art methods previously mentionned \cite{glaunes2008large,oh2024stable,wu2024shape,heo2024logistic} are compared to TS-LDDMM in \Cref{appendix: settings_classification}.
 \vspace{-1ex}
  \section{Limitations and conclusion}
  \vspace{-1ex}
  \label{sec:limitations}

  This paper proposes a feature representation method, TS-LDDMM, designed for 
shape comparison on homogeneous time series datasets. We show on a real dataset 
its ability to study, with high interpretability, the inter-individual shape 
variability. As an unsupervised approach, it is user-friendly and enables knowledge 
transfer for different supervised tasks such as classification.

Although TS-LDDMM 
is already competitive for classification, its performances can be leveraged on 
more heterogeneous datasets using a hierarchical clustering extension, which is relegated for future work. 
%   While TS-LDDMM performs well on shape-based datasets, it assumes a certain degree of homogeneity with the existence of a reference time series graph$\msg_0$.
%  Extending TS-LDDMM to more heterogeneous datasets using clustering is ongoing work.

 TS-LDDMM employs kernel computations, which require specific libraries (e.g., KeOps \cite{charlier2021kernel}) to be efficient and scalable.
 However, in our experiments, the time complexity of TS-LDDMM is comparable to that of competitors.
 It is clear that TS-LDDMM needs to be extended to handle very large datasets with high-dimensional time series (such as videos).

 Additionally, TS-LDDMM requires tuning several hyperparameters, though this is a common requirement among competitors \cite{glaunes2008large, oh2024stable, wu2024shape, heo2024logistic}.
 In future work, adaptive methods are expected to be developed to provide a more user-friendly interface.
  %\color{red} rappeler les limites de la méthodes : petite dimension à cause des kernels, tuner les hyper paramètres correctement, version adaptative}
% \paragraph*{Conclusion}
% \vspace{-1ex}
